Linear Vlasov Theory in the Shearing Sheet Approximation \\ with Application to the Magneto-Rotational Instability

### 2015-01-02 paper.tex

The salient feature of having introduced $\vvc$ is that \cref{eq:orbit-eom-vc}
takes the same form as in a uniform plasma, with the gyro-frequency vector
$\vomg$ defined in \cref{eq:gyro-frequency-vector} playing the role of the
cyclotron frequency vector $\vomc=e\vec{B}/m$. Let us therefore adopt a
cylindrical coordinate system $(\vc_\parallel,\vc_\perp,\vartheta)$ in
$\vvc$-space that is aligned with $\vomg$. In this and in the following
section, the labels $\perp$ and $\parallel$ will always refer to the direction
of $\vomg$. The steady state Vlasov equation is
\begin{equation}
  [(\vec{v} + q\Omega x\ey)\times\vec{S}]\cdot\pder{f}{\vec{v}}
  = (\vvc\times\vomg)\cdot\pder{f}{\vvc} = -\omg\pder{f}{\vartheta} = 0,
\end{equation}
where the square of the gyration frequency $\omg$ is given by
\begin{equation}
  \label{eq:gyro-frequency}
  \omg^2 = S_y^2 + S_z(S_z - q\Omega).
\end{equation}
The equilibrium distribution function is thus gyrotropic in $\vvc$-space,
i.e.\ of the form $f(\vc_\perp,\vc_\parallel)$. For this reason we will refer
to $\vvc$ as the \emph{gyrotropic velocity}.

@@

In most plasmas, electrons are much lighter than ions, and so the mass ratio
$m_e/m_i\ll1$. If we assume that the electrons are also much colder than the
ions ($T_e/T_i\ll1$), then the electrons can be modeled as a cold massless
fluid. Expanding the cold electron conductivity tensor in the mass ratio
yields
\begin{equation}
  \label{eq:cold-electron-conductivity}
  \vec{\sigma}_e = -\frac{1}{i\omega}\frac{e^2n_e}{m_e}\vec{b}\vec{b}
  + \frac{en_e}{B}\vec{b}\times\mathbf{1},
\end{equation}
where we have dropped terms of order $m_e/m_i$ and higher. The first term
evidently diverges as the mass ratio goes to zero. Because this term is
proportional to $\vec{b}\vec{b}$, the dispersion tensor becomes
block-diagonal.\footnote{We note that since
  $\mathbf{D}=-i\omega\mu_0\vec{\sigma}_e\va^2+\ldots$, it follows from the
  dispersion relation given in \cref{eq:dispersion-relation} that the parallel
  component of the (relative) electric field must be at least first order in
  $m_e/m_i$. The parallel current density would otherwise diverge, a result
  that is consistent with guiding center theory \citep[e.g.][]{Grad1961}.} In
the cold and massless electron limit, the dispersion relation thus factorizes
into $\delta{}\tilde{E}_\parallel=0$ and $\det\mathbf{D}_\perp=0$, where the
perpendicular dispersion tensor
\begin{equation}
  \label{eq:perpendicular-dispersion}
  \mathbf{D}_\perp =
  {(k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma}_i)}_\perp\va^2
  - i\omega\ome\vec{b}\times\mathbf{1}
  - 2q\Omega^2\sums\frac{n_s{}m_s}{\rho}\frac{\omcs b_z}{\omcs b_z + 2\Omega}
  (\vec{b}\times\ex)(\vec{b}\times\ex).
\end{equation}
We note that throughout this section, the labels $\perp$ and $\parallel$ refer
to the direction of $\vec{b}$, \emph{not} to the gyro-frequency vector $\vomg$
as in \cref{sec:steady-state,sec:linear-theory}. In
\cref{eq:perpendicular-dispersion}, $\vec{\sigma}_i$ is the (multi-component)
ion conductivity tensor and
\begin{equation}
  \ome = \frac{en_e B}{\rho} = \sums\frac{n_s m_s}{\rho}\omcs
\end{equation}
is the (mass density weighted) mean ion cyclotron frequency.

@@

In the limit $e/m\to\infty$ we should be able to recover the results of
\citet{Quataert2002} who analyzed the stability of a differentially rotating
plasma in the shearing sheet approximation using the so-called kinetic MHD
equations \citep{Grad1961,Kulsrud1983,Hazeltine2004}. We start by expanding
the response tensor in powers of $m/e$. Keeping mostly leading order terms,
this yields
\begin{equation}
  \mathbf{\Lambda} \approx
  -\frac{\omega^2}{\omc^2}(\mathbf{1} - \vec{e}_3\vec{e}_3)
  + \frac{i\omega}{\omc}\left[1 - \frac{(4-q)\Omega b_z}{2\omc}\right]
  \vec{e}_3\times\mathbf{1}
  - \frac{2(W-1)k_\perp^2\vt^2}{\omc^2}\vec{e}_2\vec{e}_2
  + \frac{i\zeta_0 Wk_\perp\vt}{\omc}\vec{e}_1\times\mathbf{1}
  - \zeta_0^2 W\vec{e}_3\vec{e}_3.
\end{equation}
The argument of the $W$-function is $\zeta_0=\omega/|k_\parallel|\vt$.
Expanding the anisotropy tensor $\mathbf{Q}$ yields
\begin{equation}
  \mathbf{Q} \approx \mathbf{1} - \frac{q\Omega}{2\omc b_z}
  \left[1 - \frac{(8-q)\Omega}{4\omc b_z}\right]\ey\ey.
\end{equation}

@@

\begin{figure}
  \centering\includegraphics{figures/guiding-center.pdf}
  \caption{Dispersion relation of the MRI in the guiding center limit
    ($\omc/\Omega\to\infty$) for an inclined magnetic field with $b_z=b_y$.
    Labels give the value of $\beta_z=\beta/b_z^2$. These results agree well
    with the analogous plots shown in figs.~3 and 4 of
    \citet{Quataert2002}.}\label{fig:guiding-center}
\end{figure}

@@

In this section we study modes whose wave vector is aligned with the magnetic
field, which in turn is aligned with the rotation axis. The ion response
tensor in this case is given by
\begin{equation}
  \label{eq:parallel-response}
  \mathbf{\Lambda} =
  \frac{1}{2}\sum_\pm\Lambda^{}_\pm
  (\vec{e}_1\pm i\vec{e}_2)(\vec{e}_1\mp i\vec{e}_2)
  - \frac{\omega^2}{k_z^2\vt^2} W\left(\frac{\omega}{|k_z|\vt}\right)
  \vec{e}_3\vec{e}_3,
\end{equation}
with perpendicular components
\begin{equation}
  \label{eq:circular-components}
  \Lambda_\pm = \frac{\omega}{\omega\pm\omg}
  \left[1 - W\left(\frac{\omega\pm\omg}{|k_z|\vt}\right)\right].
\end{equation}
We note that $\ex\ne\vec{e}_1$ and $\ey\ne\vec{e}_2$ in general even for
parallel modes. This is because the orthonormal basis
$(\vec{e}_1,\vec{e}_2,\vec{e}_3)$ with $\vec{e}_3=\vomg/\omg$ is taken to be
right-handed (see \cref{sec:plasma-response}), and whether or not the
gyro-frequency vector $\vomg$ defined in \cref{eq:gyro-frequency-vector} is
aligned with the angular velocity vector $\vec{\Omega}=\Omega\ez$ depends on
the sign of $S_z=\omc{}b_z+2\Omega$. In this context we also note that the
gyration frequency $\omg>0$ throughout this text.\footnote{Unless of course
  $\omg^2<0$, in which case much of our analysis is invalidated, see the
  discussion in \cref{sec:delta}.} The cyclotron frequency $\omc=eB/m$ can
have either sign, but this depends only on the sign of the charge, \emph{not}
on direction of the magnetic field.

@@

\begin{figure}
  \centering\includegraphics{figures/vf-mri-2D.pdf}
  \caption{Growth rate and wavenumber of the fastest growing MRI mode for
    $\vec{k}\bm{\parallel}\vec{B}\bm{\parallel}\vec{\Omega}$ based on a
    numerical solution of the kinetic ion, cold, massless electron shearing
    sheet dispersion relation \labelcref{eq:D-perp}. The dashed black lines
    correspond to $\beta\Omega=\omc{}b_z$, the approximate ratio of the
    gyro-viscous force to the tension force.}\label{fig:vf-mri-2D}
\end{figure}

\begin{figure}
  \centering\includegraphics{figures/vf-mri.pdf}
  \caption{Growth rate and wavenumber of the fastest growing MRI mode as
    functions of $\omc{}b_z$ at fixed $\beta$. The cusps (upper panels) and
    discontinuities (lower panels) correspond to one local minimum of
    $\gamma(k_z)$ taking over another, see \cref{fig:drel}.}\label{fig:vf-mri}
\end{figure}

\begin{figure}
  \centering\includegraphics{figures/drel.pdf}
  \caption{Plots of
    growth rate vs.\ wave number for select values of $\omc{}b_z$ and $\beta$.
    The lower panels illustrate the approach to the cold plasma dispersion
    relation as $\beta\to0$ when the tidal anisotropy is
    large.}\label{fig:drel}
\end{figure}

>>>>

The salient feature of having introduced $\vvc$ is that \cref{eq:orbit-eom-vc}
takes the same form as in a uniform plasma, with the gyro-frequency vector
$\vomg$ defined in \cref{eq:gyro-frequency-vector} playing the role of the
cyclotron frequency vector $\vomc=e\vec{B}/m$. Let us therefore adopt a
cylindrical coordinate system $(\vc_\parallel,\vc_\perp,\vartheta)$ in
$\vvc$-space that is aligned with $\vomg$. Here and in the following, $\perp$
and $\parallel$ will always be defined with respect to the direction of
$\vomg$. The steady state Vlasov equation is
\begin{equation}
  [(\vec{v} + q\Omega x\ey)\times\vec{S}]\cdot\pder{f}{\vec{v}}
  = (\vvc\times\vomg)\cdot\pder{f}{\vvc} = -\omg\pder{f}{\vartheta} = 0,
\end{equation}
where the square of the gyration frequency $\omg$ is given by
\begin{equation}
  \label{eq:gyro-frequency}
  \omg^2 = S_y^2 + S_z(S_z - q\Omega).
\end{equation}
The equilibrium distribution function is thus gyrotropic in $\vvc$-space,
i.e.\ of the form $f(\vc_\perp,\vc_\parallel)$. For this reason we will refer
to $\vvc$ as the \emph{gyrotropic velocity}.

@@

In most plasmas, electrons are much lighter than ions, and so the mass ratio
$m_e/m_i\ll1$. If we assume that the electrons are also much colder than the
ions ($T_e/T_i\ll1$), then the electrons can be modeled as a cold massless
fluid. Expanding the cold electron conductivity tensor in the mass ratio
yields
\begin{equation}
  \label{eq:cold-electron-conductivity}
  \vec{\sigma}_e = -\frac{1}{i\omega}\frac{e^2n_e}{m_e}\vec{b}\vec{b}
  + \frac{en_e}{B}\vec{b}\times\mathbf{1},
\end{equation}
where we have dropped terms of order $m_e/m_i$ and higher. The first term
evidently diverges as the mass ratio goes to zero. Because this term is
proportional to $\vec{b}\vec{b}$, the dispersion tensor becomes
block-diagonal.\footnote{We note that since
  $\mathbf{D}=-i\omega\mu_0\vec{\sigma}_e\va^2+\ldots$, it follows from the
  dispersion relation given in \cref{eq:dispersion-relation} that the parallel
  component of the (relative) electric field must be at least first order in
  $m_e/m_i$. The parallel current density would otherwise diverge, a result
  that is consistent with guiding center theory \citep[e.g.][]{Grad1961}.} In
the cold and massless electron limit, the dispersion relation thus factorizes
into $\delta{}\tilde{E}_\parallel=0$ and $\det\mathbf{D}_\perp=0$, where the
perpendicular dispersion tensor
\begin{equation}
  \label{eq:perpendicular-dispersion}
  \mathbf{D}_\perp =
  {(k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma}_i)}_\perp\va^2
  - i\omega\ome\vec{b}\times\mathbf{1}
  - 2q\Omega^2\sums\frac{n_s{}m_s}{\rho}\frac{\omcs b_z}{\omcs b_z + 2\Omega}
  (\vec{b}\times\ex)(\vec{b}\times\ex).
\end{equation}
Here, $\vec{\sigma}_i$ is the (multi-component) ion conductivity tensor and
\begin{equation}
  \ome = \frac{en_e B}{\rho} = \sums\frac{n_s m_s}{\rho}\omcs
\end{equation}
is the (mass density weighted) mean ion cyclotron frequency.

@@

In the limit $e/m\to\infty$ we should be able to recover the results of
\citet{Quataert2002} who analyzed the stability of a differentially rotating
plasma in the shearing sheet approximation using the so-called kinetic MHD
equations \citep{Grad1961,Kulsrud1983,Hazeltine2004}. We start by expanding
the response tensor in powers of $m/e$. Keeping mostly leading order terms,
this yields
\begin{equation}
  \mathbf{\Lambda} \approx
  -\frac{\omega^2}{\omc^2}(\mathbf{1} - \vec{e}_3\vec{e}_3)
  + \frac{i\omega}{\omc}\left[1 - \frac{(4-q)\Omega b_z}{2\omc}\right]
  \vec{e}_3\times\mathbf{1}
  - \frac{2(W-1)k_\perp^2\vt^2}{\omc^2}\vec{e}_2\vec{e}_2
  + \frac{i\zeta_0 Wk_\perp\vt}{\omc}\vec{e}_1\times\mathbf{1}
  - \zeta_0^2 W\vec{e}_3\vec{e}_3.
\end{equation}
The argument of the $W$-function is $\zeta_0=\omega/|k_\parallel|\vt$ with
$k_\parallel=\vec{k}\cdot\vec{b}$. Expanding the anisotropy tensor
$\mathbf{Q}$ yields
\begin{equation}
  \mathbf{Q} \approx \mathbf{1} - \frac{q\Omega}{2\omc b_z}
  \left[1 - \frac{(8-q)\Omega}{4\omc b_z}\right]\ey\ey.
\end{equation}

@@

\begin{figure}
  \centering\includegraphics{guiding-center.pdf}
  \caption{Dispersion relation of the MRI in the guiding center limit
    ($\omc/\Omega\to\infty$) for an inclined magnetic field with $b_z=b_y$.
    Labels give the value of $\beta_z=\beta/b_z^2$. These results agree well
    with the analogous plots shown in figs.~3 and 4 of
    \citet{Quataert2002}.}\label{fig:guiding-center}
\end{figure}

@@

In this section we study modes whose wave vector is aligned with the magnetic
field, which in turn is aligned with the rotation axis. The ion response
tensor in this case is given by
\begin{equation}
  \label{eq:parallel-response}
  \mathbf{\Lambda} =
  \vec{e}_3\vec{e}_3 + \frac{1}{2}\sum_\pm\Lambda^{}_\pm
  (\vec{e}_1\pm i\vec{e}_2)(\vec{e}_1\mp i\vec{e}_2),
\end{equation}
with components
\begin{equation}
  \label{eq:circular-components}
  \Lambda_\pm = \frac{\omega}{\omega\pm\omg}
  \left[1 - W\left(\frac{\omega\pm\omg}{|k_z|\vt}\right)\right].
\end{equation}
We note that $\ex\ne\vec{e}_1$ and $\ey\ne\vec{e}_2$ in general even for
parallel modes. This is because the orthonormal basis
$(\vec{e}_1,\vec{e}_2,\vec{e}_3)$ with $\vec{e}_3=\vomg/\omg$ is taken to be
right-handed (see \cref{sec:plasma-response}), and whether or not the
gyro-frequency vector $\vomg$ defined in \cref{eq:gyro-frequency-vector} is
aligned with the angular velocity vector $\vec{\Omega}=\Omega\ez$ depends on
the sign of $S_z=\omc{}b_z+2\Omega$. In this context we also note that the
gyration frequency $\omg>0$ throughout this text.\footnote{Unless of course
  $\omg^2<0$, in which case much of our analysis is invalidated, see the
  discussion in \cref{sec:delta}.} The cyclotron frequency $\omc=eB/m$ can
have either sign, but this depends only on the sign of the charge, \emph{not}
on direction of the magnetic field.

@@

\begin{figure}
  \centering\includegraphics{vf-mri-2D.pdf}
  \caption{Growth rate and wavenumber of the fastest growing MRI mode for
    $\vec{k}\bm{\parallel}\vec{B}\bm{\parallel}\vec{\Omega}$ based on a
    numerical solution of the kinetic ion, cold, massless electron shearing
    sheet dispersion relation \labelcref{eq:D-perp}. The dashed black lines
    correspond to $\beta\Omega=\omc{}b_z$, the approximate ratio of the
    gyro-viscous force to the tension force.}\label{fig:vf-mri-2D}
\end{figure}

\begin{figure}
  \centering\includegraphics{vf-mri.pdf}
  \caption{Growth rate and wavenumber of the fastest growing MRI mode as
    functions of $\omc{}b_z$ at fixed $\beta$. The cusps (upper panels) and
    discontinuities (lower panels) correspond to one local minimum of
    $\gamma(k_z)$ taking over another, see \cref{fig:drel}.}\label{fig:vf-mri}
\end{figure}

\begin{figure}
  \centering\includegraphics{drel.pdf}
  \caption{Plots of
    growth rate vs.\ wave number for select values of $\omc{}b_z$ and $\beta$.
    The lower panels illustrate the approach to the cold plasma dispersion
    relation as $\beta\to0$ when the tidal anisotropy is
    large.}\label{fig:drel}
\end{figure}
